---
title: "hca & hcaquery presentation"
author: "Maya Reed McDaniel"
date: "6/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Setup
```{r setup, message = FALSE}
library(dplyr)
library(LoomExperiment)
library(hca)
library(hcaquery)
```

## hca
- explore all available projects
```{r}
hca::projects()
```

- explore just those projects for which the organ is brain and species is Homo Sapiens
```{r}
brain_homo_sapien_filter <- hca::filters(
    genusSpecies = list(is = "Homo sapiens"),
    organ = list(is = c("brain")),
    workflow = list(is = c("optimus_v4.2.2")))

brain_homo_sapien_proj <- hca::projects(filters = brain_homo_sapien_filter)

brain_homo_sapien_proj
```

- Select a project to look at in greater depth
```{r}
project_id <- brain_homo_sapien_proj$projectId[3]
```

- What files are associated with this project
```{r}
file_filter <- hca::filters(projectId = list(is = project_id))
proj_files <- hca::files(filter = file_filter)
proj_files
```

- We want loom files for which there is a single donor (taking the two smallest)
```{r}
additional_file_filters <- hca::filters(projectId = list(is = project_id),
                                        fileFormat = list(is = "loom"),
                                        donorCount = list(is = c(1)))
proj_files_refined <- hca::files(filter = additional_file_filters,
                                 sort = "fileSize",
                                 order = "asc",
                                 size = 2)
proj_files_refined
```

- download these files
```{r}
project_file_locations <- proj_files_refined %>%
                            hca::files_download()
project_file_locations
```

## hcaquery
- let's take the smallest two `.loom` files from the Optimus workflow across all projects
```{r}
test_file_filter <- hca::filters(workflow = list(is = c("optimus_v4.2.2")),
                                 fileFormat = list(is = c("loom")))
test_file_tbl <- hca::files(filters = test_file_filter,
                            size = 2, sort = "fileSize", order = "asc")
project_file_locations <- test_file_tbl %>% hca::files_download()
```

- let's take a look at a few of there `.loom` files
```{r}
exp_1 <- LoomExperiment::import(project_file_locations[1],
                                   type = "SingleCellLoomExperiment")
metadata(exp_1)

exp_2 <- LoomExperiment::import(project_file_locations[2],
                                   type = "SingleCellLoomExperiment")
metadata(exp_2) 
```

- perhaps we could load them into the database for better querying
```{r}
hcaquery::files_to_db(test_file_tbl)
```

- description of each table
```{r}
hcaquery::available_tables()
hcaquery::table_description("genes_tbl")
hcaquery::table_description("cells_tbl")
hcaquery::table_description("assays_tbl")
hcaquery::table_description("experiment_overviews")
```

- query for file with donor species of Homo sapiens
```{r}
## example of queries on accessible db tables
hca_tbls <- hca_tables()
human_experiments <- hca_tbls$experiments_tbl %>%
    filter(donor_organism_genus_species == "Homo sapiens")

human_experiments
```
- query for specific genes within a file, returnig a `SingleCellExperiment` object
```{r}
genes <- c("Gm18956", "Gm37180", "Gm37363")
smat <- hca_file_gene_query(genes = genes,
                            file_ident = "b077b456-6362-59ea-a5b6-230dd8bc6d91")
smat
```
